openarchaeoCollaboration

### 2023-03-30 analysis/02-background.qmd
Re-wrote the third section with a more positive tone

* What is open source?

@@

* How does open source relate to open science?

@@

* Situating open archaeology in open science

The "archaeological toolbox" is a common metaphor that describes the set of physical and informational tools and processes that archaeologists rely on to achieve their goals. In applying either physical or digital methods, archaeology has a long tradition of adopting common or "off the shelf" tools to serve their own unique purposes, which are not typically unanticipated by their original designers. In other words, the apparatus of archaeology is an open system that relies on products and skillsets from outside the discipline, but which exhibit enough flexibility to be used in novel ways. For example, archaeologists are content to outsource the metallurgical skills needed to forge their trowels, and delegate the manufacture and acquisition of key electronic tools, like laptop computers and cameras, to the global supply chains and labour networks that produce consumer-grade products for purchase. However, an increasingly large number of archaeologists are taking it upon themselves to program their own software, rather than rely on generic and commercially-produced information processing systems. In fact, archaeologists are prolific computer programmers and have been coding a variety of functions and libraries that serve uniquely archaeological problems and purposes. This software is typically made to support tasks that archaeologists find uniquely meaningful, and which are not supported by mainstream technology companies. This differs from the use of purpose-specific scripts written to address specific research questions. Such work often falls under the purview of open science, since making one's scripts openly available is said to enhance the potential for reproducibility (Marwick and Schmidt).

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a desire to facilitate novel research opportunities, to make participation in scienctific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. Open science advocates have put a lot of effort into developing information infrastructures, or systems designed to facilitate the transmission of data on a global scale, but, at least among archaeologists, their transformative value has not yet been realized. For instance, numerous studies have highlighted how archaeologists are not reusing data published on open data platforms, and instead opt to communicate directly with those who created the dataset to obtain a relationship that more closely resembles a more traditional form of collaboration (Huggett 2018; Atici 2013; Faniel 2013; Sobotkova 2018; Opitz 2021). In other words, these pltforms do little to alter the professional norms and expectations that frame archaeological research, and which seem to be fundamental factors that influence the nature of archaeological research. Open data platforms may provide the capacity and the potential to enable change, but do not in themselves change the ways in which archaeologists create and value research outcomes. This must come from community buy-in, and organic extension and establishing of novel forms of collaborative commitments that reflect the ideal state of affairs. In other words, it is a social problem, not a technical one.

Similarly, archaeological software development is framed by the circumstances under which research that supports and/or motivates the effort occurs. This means that software development processes are shaped by the pressures and circumstances of being an archaeologist, such as the need to publish and to generate stable research outputs, archaeologists' tendency to move on from projects after they are 'completed' (i.e. when money runs out and people lose interest), the need to be recognized as a citable author for purposes of hiring and promotion, and the general lack of formal digital literacy among archaeologists (which either produces a digital archaeology subdiscipline, a kind of ad-hoc cobbled together system (made by the most digitally literate person in the room), or cross-disciplinary connections with non-archaeologists (spouses, childhood friends, maybe a computer scientists, rarely someone in private industry)). (Morgan 2015; Morgan and Eve 2012; Batist et al 2021; Caraher 2022; Graham 2020; maybe also Thorpe 2012 and Zorzin 2015 on economic models that frame (non-digital) archaeological research)

A lot of work has gone in to developing information infrastructures, or systems designed to facilitate the transmission of open materials on a global scale, including open data repositories, open peer-review platforms, and online social spaces meant to foster participation in new forms of scholarly communication. However, the general failure of open science to deliver on its lofty promises has led archaeologists to reflect on what openness really means and entails. In the past, open archaeology was very much wrapped up in discourse on the best tools to use, optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards (e.g., establishing proper and secure web servers, open api practices, using orcid, proper licencing, using markdown, git, R package development practices, etc) (cf. Kansa and Kansa 2014; Kintigh 2014; Roosevelt 2015; others), and this continues to be a prominent part of discourse on the development of open infrastructures (cf. Kansa on open identifiers; others). However, more recent reflections have attempted to re-frame research practices, particularly those corresponding with "data work", as situated experiences and associated with a continuum of decisions made within pragmatic contexts of action. Archaeological research is thus framed as inherently collaborative, and the materials and media that archaeologists rely on to do their research are framed as discursive entities, or communicatve media that bridge a gap of experience or understanding. Bridging this gap entails an act of reconciliation between different pragmatic experiences, and the technical infrastructures we have set up to handle these acts of communication do not adequately account for the range of factors that must be considered to successfully pass this reconciliation.

>>>>

## What is open source?

@@

## How does open source relate to open science?

@@

## Open archaeology at a crossroads

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a desire to facilitate novel research opportunities, to make participation in scienctific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. Open science advocates, including archaeologists, have put a lot of effort into developing information infrastructures, or systems designed to facilitate the transmission of data on a global scale. And as with open science in general, open archaeology has been wrapped up in discourse on the best tools to use, optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards (cf. Kansa and Kansa 2014; Kintigh 2014; Roosevelt 2015). This continues to be a prominent part of discourse on the development of open infrastructures (cf. Kansa 2022 on unique  identifiers; something about aDNA pipelines; Hacıgüzeller 2021 on CIDOC-CRM (though she is actually very critical); Huvila and his team on formal paradata), but more recent reflections on the use of these infrastructures recognize underlying and systemtic commitments and assumptions that inform how these systems are used and valued by various stakeholders. For instance, the pressures and circumstances of being an archaeologist, such as the need to publish stable research outputs, and archaeologists' tendency to move on from projects after they are completed, contribute towards shaping data sharing platforms as delegated archival services, whereupon the obligation to care for data over the long-term is passed on to others. In response to criticisms that highlight how these systems add little practical downstream value (Huggett 2018; Sobotkova 2018), open archaeology advocates have begun to prioritize the role that information infrastructures may play in contexts of data reuse, namely by implementing the FAIR principles in order to ensure that data are findable, accessible, interoperable, and reusable (Kansa 2023 in that recent special issue in Advances in Archaeological Practice; one or two additional sources). This example is matched by similar efforts to raise the value proposition of using open science tools and systems, such as through an emphasis on enhancing R package development and documentation skills so that code may be run and evaluated in alternative computing environments (cite Marwick and/or Roe), and [another example].

Additionally, archaeologists are becoming more comfortable showcasing and actively discussing undervalued research ouputs such as software and datasets, which are reviewed and published by venues like Peer Community In, the Journal of Archaeological Data, and the Journal of Open Source Software. Professional blogs, social media and podcasts are also gaining traction as proper venues for disseminating one's work. This is consistent with archaeology's historical willingness to experiment with new methods and media and its casual attitude towards scholarly communication. Archaeologists' embrace of methodological plurality and their positions as interdisciplinary scholars scattered across institutional boundaries has prepared them for this kind of pragmatic, improvisatory approach to scholarly communication via the web. There is enormous potential for archaeologists to reinvent novel forms of scholarly communication and extend a new set of collaborative commitments that go beyond those typically recognized by scholarly institutions, and which correspond with open science principles. But this depends on archaeologists' willingness to continue experimenting and on the community's willingness to buy into novel value regimes.
